---
title: "STAT656_Project"
author: "Moein-Samira-Homa"
date: "10/8/2020"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(caret)
library(statsr)
library(data.table)
library(stats)
library(openintro)
```

#Read the data

```{r load data}
rm(list = ls())
covid_data = fread("/Volumes/GoogleDrive/Shared drives/STAT656_Project/Raw_Data/COVID_by_County_Raw.csv")
icu_data = fread("/Volumes/GoogleDrive/Shared drives/STAT656_Project/Raw_Data/ICU_bed_by_county_Raw.csv")
tot_pop_data = fread("/Volumes/GoogleDrive/Shared drives/STAT656_Project/Raw_Data/POP_Data_Race_Raw.csv")
health_data = fread("/Volumes/GoogleDrive/Shared drives/STAT656_Project/Raw_Data/us_county_sociohealth_data_Raw.csv")
```

#Extract Positive case rates and death case rates

```{r filter_covid}
covid_by_county <- covid_data %>% group_by(County, State) %>% summarise(Positive_Cases = max(Cases), Deaths = max(Deaths))
```

#Demographic data
Since Covid has higher effects on older people, the demographic data is divided into two groups of people: older ones and young ones.
It seems that, based on the literature and news, white and non-white race might have an impact on Covid positive rates and death rates. So, they are also imported to our datasets.
Finally, all data cleaned so far is merged in a table based on the mutual state and county names.

```{r}
elder_pop <- tot_pop_data %>% group_by(County, State) %>% filter(AGEGRP >= 14) %>% summarise(Elderly_pop = sum(TOT_POP))
yonger_pop <- tot_pop_data %>% group_by(County, State) %>% filter(AGEGRP < 14 & AGEGRP > 0) %>% summarise(Young_pop = sum(TOT_POP))

race_pop <- tot_pop_data %>% group_by(County, State) %>% filter(AGEGRP == 0) %>% select(WHITE, NON_WHITE, County, State)

population_data = merge(elder_pop, yonger_pop, by = c("County", "State"), all.x = TRUE)
population_data = merge(population_data, race_pop, by = c("County", "State"), all.x = TRUE)
population_data = merge(covid_by_county, population_data, by = c("County", "State"), all.x = TRUE)
merged_covid_data = merge(population_data, icu_data, by = c("County", "State"), all.x = TRUE)
merged_covid_data = merge(merged_covid_data, health_data, by = c("County", "State"), all.x = TRUE)
```

Finally, all data cleaned so far is merged in a table based on the mutual state and county names.

```{r}

merged_covid_data$POP_TOTAL <- merged_covid_data$WHITE + merged_covid_data$NON_WHITE
merged_covid_data$POP_DENSITY <- merged_covid_data$POP_TOTAL / merged_covid_data$area_sqmi
merged_covid_data$Positive_Case_Rate <- merged_covid_data$Positive_Cases / merged_covid_data$POP_TOTAL
merged_covid_data$Death_Rate <- merged_covid_data$Deaths / merged_covid_data$POP_TOTAL

final_data = merged_covid_data
```

#Plot Missing Data

```{r plot_missing_values}
ggplot_missing <- function(x){
  require(reshape2)
  require(ggplot2)
  x %>% 
    is.na %>%
    melt %>%
    ggplot(data = .,
           aes(x = Var2,
               y = Var1)) +
    geom_raster(aes(fill = value)) +
    scale_fill_grey(name = "",
                    labels = c("Present","Missing")) +
    theme_minimal() + 
    theme(axis.text.x  = element_text(angle=45, vjust=0.5)) + 
    labs(x = "Variables in Dataset",
         y = "Rows / observations")
}

ggplot_missing(final_data)
```

#Save the file

```{r save_data}
fwrite(final_data, "Final_Covid_Data_Raw.csv")
```

